Hyper-linked Communications: WebRTC enabled asynchronous collaboration

### 2016-10-29 artigo/paper.tex
Submission to ICC 2017

\documentclass[conference,compsoc,a4paper]{IEEEtran}
\usepackage{geometry}
 \geometry{
 a4paper,
 total={170mm,257mm},
 left=20mm,
 top=20mm,
 }

\usepackage[backend=bibtex]{biblatex}
\usepackage{array}
\usepackage[utf8]{inputenc}   % <<<<< Linux
\usepackage[nomain,acronym,xindy,toc]{glossaries} 
\usepackage{graphicx}
\usepackage{tabularx} 
\usepackage{float}

\usepackage{color, colortbl}
\definecolor{Gray}{gray}{0.9}
%Includes "References" in the table of contents
%\usepackage[nottoc]{tocbibind}

\bibliography{references} % or

@@

\begin{abstract}
The Hyper-linked communications concept applies much of the hypermedia concepts, widely used on Web content. This paradigm allows to synchronize, structure and navigate communication content integrated into voice and video calls.

Voice and image together can express emotions like no other medium can. With hypermedia concepts, we can add more value to conference calls.

\emph{WebRTC} technology allows real time communications between web browsers without the need to install additional software. The nature of web browser applications already follows the hypermedia concept, which makes \emph{WebRTC} the ideal technology to apply the hyper-linked communications concepts.
The web browser platform provides an abstraction layer that makes it possible to create applications that run independently from the operating system.
The native support for \emph{WebRTC} in operating systems extends its usage to outside the web browser, allowing for the exploration of functionalities for which web browsers provide poor support, such as video recording and massive information storage.

Our goal was the development of an application targeted to the web platform, resorting to \emph{WebRTC}, that leveraged the hyper-linked communications by providing a conference environment enriched with multiple media types, collaborative text editors, time annotations, instant messaging and a mechanism to superimpose hyper-content to video.
\end{abstract}

@@

\subsection{Background}
\label{section:background}

	As communications technologies appeared, we adapted the way we communicate. The purpose of this project is not the replacement of the current video and audio communications, but to enrich them with hyper-media content and make them a more natural and easy to learn process.

	With the advent of WebRTC and its successive integration with web browsers, it became possible to develop video conference web applications without plugins, this presents a range of possibilities on what can be implemented using already existing web technologies.

    Furthermore, real time communication applications can make a significant difference on business, education and health sectors by providing tools for developing teaching and learning online, teamworking and socializing web applications.

\subsection{Proposed Solution}
\label{section:proposed}

	Our goal in this project is to develop an application targeted to the web platform, resorting to \gls{WebRTC}, that leverages the hyper-linked communications by providing a video conference environment enriched with interactive and non-interactive discrete media types such as images, subtitles, forms and all types of content that can be added using \gls{HTML}5, \gls{CSS}3 and \emph{JavaScript} including continuous media types such as video, music and animations.

	One of the key features of this project is the ability to navigate in time in order to reproduce the conversation again or introduce hyper-content to it such as time annotations, interactive lists of topics and subtitles. In this context we also provide a simpler method for creating and synchronizing hyper-content using \gls{QR} codes.

	In addition to this conference environment, which provides different functionalities than traditional conference environments such as \emph{Skype} and \emph{Google Hangouts}, we also enable a collaborative text editor and a chat that supports sending time hyper-links and files to conference participants.

	Furthermore, another relevant feature is the possibility to compose multiple video streams into a single one, which enables adding more users to conference rooms without impacting on clients performance. Users can change to individual streams on demand or automatically to the talking users.

\subsection{Thesis Contribution}
\label{section:contribution}

Making it clear, this project aims to complement current audio, text and video communications in order to create rich and collaborative interfaces with the ability to add more content on a future time (\emph{e.g.} creating time annotations for improving content search) in order to increase its value. It is also important to highlight another goal of this project which is the ability to navigate in time by rewinding communications, fast-forward and jump to certain points.

	We have presented an architecture that can meet our goals, implemented the respective prototype and tested it with real users and performance benchmarks.

	According to Martin Geddes, the quality of the interaction worsens as the number of users increase\cite{geddes}. In our testing phases we will quantify and qualify the impact of increasing users on the interface and performance of our prototype.

	All the problems faced during the development and limitations were reported on the thesis so that a future project better then ours can be easily and better developed.

\subsection{Outline}

This rest of this document is structured as follows:

\begin{itemize}

\item \textbf{Chapter \ref{chapter:relatedwork}} describes the previous work in the field.
\item \textbf{Chapter \ref{chapter:architecture}} describes the system requirements and the architecture for an Web Application that fulfills the goals of this thesis.
\item \textbf{Chapter \ref{chapter:implementation}} describes the implementation of our Web Application and the technologies chosen.
\item \textbf{Chapter \ref{chapter:evaluation}} presents the evaluation tests performed and the corresponding results.
\item \textbf{Chapter \ref{chapter:conclusion}} summarizes the work developed and proposes future work.
\end{itemize}

\section{Related Work}
\label{chapter:relatedwork}

One way to overcome the \gls{IPv4} address scarcity problem \cite{ipv4} was the development of a mechanism that groups multiple address into a single one, the machine that is assigned that address is then responsible for redirecting messages to members of its group using their private addresses, each connection in the private network is identified publicly by the same \gls{IP} address with a different port. This technique is known as \gls{NAT}\cite{rfc3489}.

\gls{NAT}s weaknesses are being exposing at the application layer, namely impacting applications that require direct communications between two private networks.

In order to implement an hyper-linked communication solution, several design decisions had to be made. The limitations imposed by the \emph{Internet's} structure, its protocols and a browser's capabilities are key factors to consider when implementing a solution that allows bi-directional communications, interactive media and collaboration environment.

Due to the use of \gls{NAT}, bi-directional communications between clients have different needs from request-response based communications between clients and servers.
This lead to the appearance of mechanisms such as \gls{STUN}, \gls{TURN} and \gls{ICE}, in order to bypass the limits imposed by \gls{NAT}.

\gls{STUN}, \gls{TURN} and \gls{ICE} \cite{natvoip} servers are a possible solution to overcome \gls{NAT}'s exposed weaknesses to applications that require communications between two private networks.

When connection is established, \gls{WebRTC} came to simplify how audio and video are transmitted through web browsers. \gls{WebRTC} is an open source technology that defines a collection of standard protocols and \emph{JavaScript} \gls{API}s for web browser based real time communications without installing any additional application or plug-in.

\gls{WebRTC} uses \gls{SDP} \cite{rfc4566} to define peer connection properties such as types of supported media, \emph{codecs}, protocols used and network information. An \gls{SDP} offer describes to other peers the expected type of communication and its details, such as used transport protocols, codecs, security and other.

Some operating systems such as \emph{Android}, \emph{iOS}, \emph{Linux}, \emph{OSX} and \emph{Windows} implement native \gls{WebRTC} libraries, extending the usage of \gls{WebRTC} to applications outside the web browser. This native support can help to implement applications that record video and audio streams for further playback.

However, \gls{WebRTC} by itself does not define how users get to know each other nor how information flows between users. For this reason, we have studied the multiple ways we could implement this \emph{get-to-know} mechanism which is known as signaling protocol.

  Signaling is the process by which applications exchange connection information about peers and servers, their capabilities and meta-data.
  In particular, \gls{WebRTC} does not implement signaling, as different applications may require different protocols and there is no single answer that fits all problems.
  As a consequence, multiple options are available for filling the missing \gls{WebRTC}'s signaling component, which can be performed using \gls{SIP}\cite{rfc3261}, \gls{XMPP}, \emph{WebSockets}, \emph{Socket.io}\footnote{\url{http://socket.io/}(accessed June 1, 2015).}, \gls{SigOfly}\cite{sigofly} or by implementing a custom protocol.

  One of \gls{WebRTC} signaling's requisites is bi-directional communication. The \emph{WebSocket} protocol allows bi-directional communications over a full-duplex socket channel \cite{rfc6455}, by other words it supports sending and receiving data simultaneously.

With the communications establishment issue solved, we had to discuss the different types of media and what can be done with each kind in order to increase the value of communications among users. In this context, we have studied solutions and libraries that allow us to implement our prototype with time manipulation features, collaborative text edition, record and playback interactive video.

	\emph{Hypermedia} concept brings the possibility to organize and overlay multimedia elements into a nonlinear linear structure holding the promise of future technology and features. Languages such as \gls{SMIL}~\cite{hyval}, \emph{HyVAL}~\cite{hyval} and \gls{HTML} can be used to implement the \emph{Hypermedia} concept. Two examples of applications that captured our attention were \emph{HyperCafe}~\cite{hypercafe} and \emph{Hyper-Hitchcock}~\cite{hitchcock} which explored interactive video features.

  Using technologies that relies only on web standards, like \gls{CSS}, \gls{HTML}5, \emph{JavaScript} and \gls{SVG}, will make possible to develop an application that applies the hypermedia concept with the advantage of being compatible with a greater amount of web browsers.

\subsection{Extending collaboration tools with time manipulation}
\label{collab}

Real time collaboration applications have become a huge help on team tasks, providing a great boost on business, research and investigation velocity.

 Our first concern on real time collaboration applications is the data storage and representation. Storing multimedia content on a web client is not a viable solution because the local storage is limited to at most five megabytes per origin. Additional servers will be required to process and record the large amount of data generated by audio and video streaming.

\gls{KMS} supports streaming over \emph{WebRTC}. Another important component of \gls{KMS} is \emph{Kurento Repository}, which supports recording and playing directly from \emph{MongoDB}. That is important for providing a scalable media storage.

     \gls{OT} technology was originally developed for consistency maintenance and concurrency control over distributed objects.
     \gls{OT} algorithms are mainly used in collaborative applications such as distributed document edition.

	Among mutiple \gls{OT} platforms and libraries we present \emph{ShareJS}\footnote{\url{http://sharejs.org/}(accessed June 2, 2015).}, \emph{TogetherJS}\footnote{\url{http://togetherjs.com/}(accessed June 2, 2015).}, \emph{Goodow}\footnote{\url{http://realtimeplayground.goodow.com/}(accessed June 2, 2015).}, \emph{Etherpad Lite}\footnote{\url{https://github.com/ether/etherpad-lite}(Accessed 20 March 2016)} and \emph{otJS}\footnote{\url{http://operational-transformation.github.io}(accessed March 10, 2016)}.

	\emph{otJS} is a \emph{JavaScript} library that only implements operation transformations over plain text on the client side and requires implementing the content's persistent storage. Besides this drawback, this library is very flexible because it is not tied to a specific database or server side technology.

@@

Taking into account the goals of this project and all the technology presented so far. Our proposal is the development of a web application that provides communication and collaboration features in real time.

\subsection{Requirements}
In a general way our system's goal is to provide a multi-party video and audio conference environment that supports chat, time manipulation, collaborative text edition and hyper-content creation.

%RP isto pode ser muito expandido. Podes começar por descrever os goals (mais alto nível): audio e video conferência multi-party, possibilidade de rever o que foi dito antes, chat, etc etc.
% Só depois traduzes isso em requisitos, tanto funcionais como não funcionais
% Depois apresentas a arquitectura, fazendo um paralelo entre os módulo e escolhas realizadas e quais os requisitos que cada um deles satisfaz.

For our system, our application must provide: a simple way to send instant text messages to the conference participants, ability to recording and playback recorded video including all the hyper-content displayed at that time, mixing multiple user streams into a single stream, create annotations associated to a specified time, allow users to superimpose hyper-content to video given a range of time, search every objects related to a conference room such as hyper-content, time annotations and users, share files and time links among users, sound detection for showing the current speaker, provide a collaborative text edition tool, interpret \gls{QR} codes in order to ease content creation and support database replication.
%RP capacidade de lidar com NAT

	Moreover we allow clients to discover chat rooms and other clients by navigating on the web pages provided by our web server. In addition users can create rooms for multi-party audio and video communication communication which is achieved by using \gls{WebRTC}'s \emph{PeerConnection}.
        %RP falas em peerconnection, mas na figura 3.1 apenas há um cliente, o que obviamente impede mostrar a comunicação entre peers.
        %RP acho que podes fazer outro diagrama com a comunicação entre os vários componentes, onde apareceriam vários clientes. A aí indicas o tipo de ligação entre cada par de componentes (HTTP, PeerConnectino, websocket, MediaStream ...)

\subsection{Modules}
In this section we present the several modules that were designed in order com fulfill the set requirements.
	Figure~\ref{fig:modules} presents the structure of our system which was divided into six modules:

@@

\begin{itemize}

\item \textbf{Application module} - responsible for providing information about the relevant modules (\emph{NAT Traversal} and \emph{Signaling}) and user interface to the \emph{Client} in the form of web pages in \gls{HTML} and \emph{JavaScript} libraries through \gls{HTTP}.

\item \textbf{Signaling module} - responsible for \emph{Client} and \emph{Stream} coordination which will be performed using \emph{WebSockets}.

 \item \textbf{NAT Traversal module} - \gls{STUN} and \gls{TURN} techniques used by \emph{Client} and \emph{Stream} modules during the \emph{Signaling} phase which ends by establishing the connection between them.

 \item \textbf{Stream module} - responsible to deliver and receive multimedia content from the \emph{Client} using \gls{WebRTC}.

 \item \textbf{Storage module} - provides two main functionalities: store the model information and media recorded. This is the single module responsible for persistent storage. It stores user and communication data as well as all the data required among user sessions. It is also use to store all the communication streams, so that they can be viewed later.

 \item \textbf{Client module} - responsible for the interaction with the user.

\end{itemize}

%RP Acho que podes escrever muito mais sobre cada componente. Isto está muito curto. Fala de algumas das funções que proporcionam. Vê o que fiz com o Storage Module

%RP também podes fazer um diagrama com vários utilizadores e mostrar quais os streams que vão de e para cada um (mostrar a mistura de som, video, etc).

\subsection{Implementation Proposal}
The infrastructure is composed by: web server, stream server, signaling server, database and video repository.
%RP tens da fazer o paralelo entre isto e os componentes apresentados na secção anterior. O mesmo com a imagem, que parece não estar relacionada. Podes colocar os elementos na mesma posição, ou dentro das caixas da imagem 3.1

\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{figures/infrastructure.pdf}
	\caption{System Infrastructure}
\end{figure}

In order to simplify our solution, we propose that the \emph{Application}, \emph{Signaling}, \emph{Stream} and \emph{Storage} modules be implemented within the same server application, so it could be easy to deploy as a single image. To this set of modules, we often call \emph{backend}.

	\subsubsection{Security and authorization}

Having established that the \emph{backend} modules are placed in the same machine, that helps controlling which resources the client has permission to access as those modules are seen as a private network.
%RP na mesma ``server application'' ou ``machine''? São coisa diferentes! Apresentas vantages de instalar tudo na mesma máquina mas não falas das desvantagens. Menor escalabilidade parece-me ser logo uma delas. Na figura 3.2 aparecem várias máquinas!

To qualify the above we provide public access to \gls{HTTP} server ports, maintaining the access to other components restricted through firewall rules.
%RP qualify?
%RP Imagem com firewall e estes portos? Mas isto já me parece muito detalhe de implementação.
In relation to the database there is no directly access from the outside. All the database information is accessed via our application server which validates the permissions of users on our system.

On the other hand, the access to our streaming servers is also restricted, but clients can connect to them after concluding the signaling phase. This signaling phase may or not proceed in function of the client access permissions. For example, if a user is trying to access a private conference room that he is not a member of nor has an invitation link for, the signaling server refuses to start the signaling phase and the user cannot access the streaming server.

The placement of our streaming servers inside a \gls{NAT} also has an important role with respect to external misuse prevention. Otherwise, placing our streaming servers could allow external clients to perform their own signaling protocol and, as a consequence, use our infrastructure without our consent.

\subsubsection{Client connections}
	Although the delegation of processing work to clients can improve our system's scalability, we are concerned about using the least resources possible on the client side, as huge resource consumption may drain battery very fast or may even be impossible to run on mobile devices. We are aware that streaming video from clients is already a very intensive task which we cannot avoid but can improve by delegating the most intensive tasks to our servers.

	In this context, with a centralized approach, each client must only have one \emph{PeerConnection} to our streaming server and content shown to them is changed on demand either being it an individual or a composite view. Otherwise clients could follow a peer-to-peer connection which would result on maintaining more connections and performing the composition of videos on client side.

	The composition of streams on client side is performed by receiving streams with the best quality possible but, due to undersizing the video of clients into a smaller region, this would result on wasting bandwidth on a quality that is not needed.

	Although the peer-to-peer approach could be used on our system, we conclude that we need to record the video on our streaming server because web clients have a very limited storage and peer disconnections may result on recorded video loss.

	The same can be concluded to instant message delivery, each client must have only one \emph{WebSocket} connection to the application server which consequently relays the messages to other users.

	Relaying instant messages from clients through the our web servers is easier if all clients are connected to the same server because all messages can be directly delivered without sending messages across multiple web servers.

	In the context of this thesis, we will not implement sending messages across web servers but, in order to allow our system to scale, we will consider that all conference participants are connected to the same server and our system is scaled by having conference rooms distributed across different servers.

\subsubsection{Software choices}

Furthermore, we have taken into account the compatibility between the streaming server, database and the operation transformation solutions, in order choose the appropriate framework to implement our web server.
%RP esta secção devia começar aqui, a explicar a escolha das tecnologias/produtos a usar em cada componente. O que tens antes são pormenores que só faz sentido discutir no final. Tens de ir sempre do mais geral para o mais detalhado.

We have decided that our solution must use \gls{KMS}. Our web server could be implemented easily with \emph{NodeJS} or \emph{Java} due to the fact \gls{KMS} provides clients for both technologies. But others could also be used as \gls{KMS} also exposes their \gls{API} via \emph{WebSockets}.

Due to the fact we are going to use \gls{KMS} as streaming server solution, we could choose \emph{NodeJS} or any \emph{Java} based web framework for implementing our web application server. We have decided to implement our web server with the \emph{PlayFramework}\footnote{\url{https://www.playframework.com/}(acessed March 25, 2016)} using \emph{Java} because of our previous experience with it.

By default, \emph{Kurento Repository} is implemented over \emph{MongoDB}, for convenience our storage model will also use the same database.

Importantly, for the collaborative text editor, we have chosen \emph{OT.js} due to its server and storage implementation choice independence.

For the \emph{NAT Traversal} module a public \gls{STUN} server can be used for testing our solution. Nevertheless, we recognize that for a production environment we would need to maintain our own \gls{TURN} servers in order to ensure connectivity to all clients.

>>>>

%\documentclass[conference,compsoc,a4paper]{IEEEtran}
\documentclass[10pt,conference]{IEEEtran}

\usepackage{array}
\usepackage[utf8]{inputenc}   % <<<<< Linux
\usepackage[nomain,acronym,xindy,toc]{glossaries} 
\usepackage{tabularx} 
\usepackage{float}

\usepackage{cite}    
\usepackage{graphicx,flafter}
\graphicspath{{../tese2/} {.}}

\usepackage{color, colortbl}
\definecolor{Gray}{gray}{0.9}

\usepackage{url}

@@

\begin{abstract}
The Hyper-linked communications concept applies much of the hypermedia concepts, widely used on Web content, to the communications realm.
This paradigm allows to synchronize, structure and navigate communication content integrated into voice and video calls.
Voice and image together can express emotions like no other medium can. 
With hypermedia concepts, we can add even more value to conference calls.

This paper describes the design, implementation and evaluation of a hyper-media communication platform targeting the web platform and making use of \emph{WebRTC} technology.
Our prototype provides a hyper-linked, non-linear, multi-party video conferencing platform with support for multiple media types, collaborative text editor, time annotations, instant messaging and a mechanism to superimpose hyper-media to video.

%Our prototype was built using a hybrid peer-to-peer and client-server architecture, resorting to open source software.
Usability tests using 20 users evaluated our prototype positively: 100\% considered it innovative and 95\% recommended its use. 
From a deployment point of view, our performance tests showed that our solution was expensive to scale up, even though the cost could be significantly lowered by disabling some of the advanced functionality.

\end{abstract}

@@

As communications technologies evolved, we adapted the way we communicate. 
Multi-party video conference is today a commonly used tool in the enterprise, academia and personal live.
The \gls{WWW} has made us familiar with the concept of text hyper-media, in the form of hyper-links.
\gls{IPTV} and video hosting platforms, such as Youtube, made non-linear video watching a common practice.
The latter also use the concept of hyper-links superimposed over videos.
However, these three concepts, video-conferencing (which is real-time), non-linear video watching and hyper-media, have largely remained independent concepts.

This paper describes our efforts to develop a prototype that combines these three concepts, providing a video-conferencing web application that allows users to navigate back and forth in time to catch up on what was said, while the video-conferencing is going on.
We developed a collaboration application, targeted at the web platform resorting to \gls{WebRTC}, that leverages the hyper-linked communications concept by providing a video conference environment enriched with interactive and non-interactive discrete media types such as images, subtitles, forms and all types of content that can be added using \gls{HTML}5, \gls{CSS}3 and \emph{JavaScript}.
Our collaboration application also includes a collaborative text editor and a chat that supports sending time hyper-links and files to conference participants.

One of the key features of this project is the ability to navigate in time in order to reproduce the conversation again.
Users can use a time bar to navigate in time: rewinding communications, fast-forward or jump to a certain point in time.
Time hyper-links also enable accessing the collaboration (video, chat, text editor, etc) at a given instance in time.
Users are capable of enriching the video content, in real-time or at a latter time, by adding content to it, such as timed annotations, interactive lists of topics and subtitles, thus also improving content search abilities.
In this context, we also provide a method for real-time creation and synchronizing of hyper-linked content using \gls{QR} codes.

We envision the use of our collaboration platform in many situations, such as: virtual classroom, where students can interact with the professor during class and later review the subjects, while the latter can enrich the recording with extra materials and annotation after the lecture; recording meetings, where the video stream can be annotated and bookmarked with the agenda points, and support documents added.

The remainder of this paper is organized as follows.
Section \ref{chapter:relatedwork} provides a background on the field.
Section \ref{chapter:architecture} describes the requirements we defined for our system and architecture for fulfilling them.
The choices performed for the implementation of our prototype are presented in Section \ref{chapter:implementation}.
In Section \ref{chapter:evaluation} we present the evaluation tests performed and their results.
Section \ref{chapter:conclusion} summarizes the work developed and proposes future work.

\section{Background}
\label{chapter:relatedwork}

\emph{Hypertext} is a type of text that provides links to texts or other types of content, these links are known by \emph{hyperlinks}. 
\emph{Hypermedia} is an evolution of \emph{hypertext}, it includes audio, images, text and video. 
\emph{Hypermedia} concept brings the possibility to organize and overlay multimedia elements into a nonlinear structure~\cite{geddes}.

Some \gls{MPEG} implementations added hypermedia information to empty space present on \gls{MPEG} frames in order to provide interactive television, modifying the \gls{MPEG} encoder and decoder in order to handle hypermedia content~\cite{embedded}.

Hyper-video is a kind of video that contains links to any kind of hypermedia, including links to skip part of it. An example of hypermedia application could be a search engine over hypermedia content, like subtitles, in order to jump to a specific time in a video or audio track.
\emph{HyperCafe} was an experimental project to expose hyper-video concepts that consisted of an interactive film that enabled switching between different conversations taking place inside a cafe~\cite{hypercafe}.

Detail-on-demand is a subset of hyper-video that allow us to obtain additional information about something that appears along the video, like obtaining information about a painting that appears in a particular segment. 
\emph{Hyper-Hitchcock} is an editor and player of detail-on-demand video~\cite{hitchcock}.

\emph{HyVAL}~\cite{hyval} is an \gls{XML} based language that was proposed for modeling composition, synchronization and interaction of hypermedia. 
HyVAL defines defines video structure, internal video and external media objects. 
\emph{HyVAL} uses a primary video stream, around which all other elements are organized and synchronized.
\emph{HyVAL}'s video structure object defines a structure derived from traditional video, which divides video into segments, scenes, shots and frames hierarchically. 
  This approach is quite restrictive if we want to apply hyper-video concepts to videos that do not follow this structure. 
%External media objects are linked by primary video, those objects can represent other videos, images, text, animation and sound.
  %RP não percebo bem a última frase. O que é primary video?

  \gls{SMIL}~\cite{smil} was introduced to describe temporal behavior of multimedia content, in particular, it could be used to overlay subtitles on films. 
With \gls{SMIL} it is possible to synchronize multiple videos, either in parallel or in sequence, reproduce a different audio track, overlay user interface elements with hyper-links, among multiple other features.
  %RP synchronize multiple sections of video? vários videos apresentados em simultâneo?

 % \gls{SMIL} is an \gls{XML} based language that defines twelve modules: \emph{Animation}, \emph{Content Control}, \emph{Layout}, \emph{Linking}, \emph{Media Objects}, \emph{SmilText}, \emph{Meta Information}, \emph{Structure}, \emph{Timing}, \emph{Time Manipulations}, \emph{State} and \emph{Transitions}.

%The \gls{DOM} is a standard \gls{API} that allows easy management of documents that are organized in a tree structure, by providing \gls{CRUD} operations over its elements and their attributes. \gls{DOM} makes it easy to inter-operate between imperative and declarative programming languages~\cite{dom}.
%RP não arranjas uma referência para o dom?

%Like \gls{DOM}, \gls{SMIL} \gls{DOM} is an \gls{API} for \gls{SMIL} documents. Allowing \gls{CRUD} operations over \gls{SMIL} documents is an important feature for extending \gls{SMIL} capabilities, for example for creating non-linear animations and triggering external events like \emph{JavaScript} functions.

%  \gls{SMIL} fits our goals for creating a multimedia rich hyper-call, but it lacks on browser compatibility. Ambulant \cite{ambulant} was one of the SMIL players that were developed for browsers, although this player implements most of \gls{SMIL} 3.0 \cite{smil3} specifications, it needs to be installed on browsers as a plug-in.

%  SmillingWeb \cite{smillingweb} attempts to implement a cross platform multimedia player designed for \gls{SMIL} 3.0 presentations with \emph{JavaScript} and \emph{jQuery} which, unlike \cite{ambulant}, does not require a plug-in to be installed and should not have incompatibility issues. 
%  SmillingWeb already takes advantage of \gls{HTML}5 and \gls{CSS}3.
%  It takes into account unsupported web browsers through the use of \emph{Modernizr}\footnote{\url{http://modernizr.com/}(accessed June 2, 2015).}, a simple \emph{JavaScript} library that may require plug-ins if new features are not supported.  
%  But SmillingWeb just implements a subset of \gls{SMIL} 3.0 and their scheduler engine loads the \gls{SMIL} file only once, which could raise problems when dealing with \gls{SMIL} changes due to real time communications.
 % Another problem with SmillingWeb is pre-loading and playing elements at the correct interval of time, which is not always possible due to high latency networks leading to  pauses during playback.

 % An alternative to \gls{SMIL}'s \emph{Layout} module is to use \gls{HTML} which is a markup language based on \gls{XML} that is used for creating web pages. \gls{HTML} alone is a very poor language when we are focused on visual appealing and interactive web pages. Languages like \gls{CSS} and \emph{JavaScript} are typically used along with \gls{HTML} for improving the interaction and appearance of a web page.

  With the emergence of \gls{HTML}5, tags like \emph{video}, \emph{audio} and \emph{track} allow us to play video with multiple \emph{codecs}, audio and subtitles in \gls{WebVTT} format.
Another important tag is \emph{canvas} that allows drawing graphics with \emph{JavaScript} on a rectangle within a web page.

%  For example, with \gls{API}s like WebGL\footnote{\url{http://khronos.org/webgl/}(accessed June 2, 2015).}, it is now possible to manipulate a three dimensional environment in the context of a hyper-call. Another example would be a collaborative spreadsheet using \gls{WebRTC}. With this, hyper-calls are not limited to only audio, image, text and video, but also interaction with complex graphical user interfaces that changes over time.

  \gls{SVG} is an \gls{XML} based format that incorporates the animation module of \gls{SMIL}. Currently, \gls{SVG} allows adding movement and animating attributes of elements. When embedded on \gls{HTML}5, it allows dynamic changes to inner content in real time through the \gls{DOM} \gls{API}. Besides that, it also allows calling \emph{JavaScript} functions on events such as animation end, mouse over and mouse click.

\gls{WebRTC} is an open source technology that defines a collection of standard protocols and \emph{JavaScript} \gls{API}s for web browser based real time communications without installing any additional application or plug-in. 
\gls{WebRTC} uses \gls{SDP} \cite{rfc4566} to define peer connection properties such as types of supported media, \emph{codecs}, protocols used and network information. An \gls{SDP} offer describes to other peers the expected type of communication and its details, such as used transport protocols, codecs, security and other.

%Some operating systems such as \emph{Android}, \emph{iOS}, \emph{Linux}, \emph{OSX} and \emph{Windows} implement native \gls{WebRTC} libraries, extending the usage of \gls{WebRTC} to applications outside the web browser. This native support can help to implement applications that record video and audio streams for further playback.

However, \gls{WebRTC} by itself does not define how users get to know each other nor how information flows between users. 
  Signaling is the process by which applications exchange connection information about peers and servers, their capabilities and meta-data.
  In particular, \gls{WebRTC} does not implement signaling, as different applications may require different protocols and there is no single answer that fits all problems.
  As a consequence, multiple options are available for filling the missing \gls{WebRTC}'s signaling component, which can be performed using \gls{SIP}\cite{rfc3261}, \gls{XMPP}, \emph{WebSockets}, \emph{Socket.io}\footnote{\url{http://socket.io/}(accessed June 1, 2015).}, \gls{SigOfly}\cite{sigofly} or by implementing a custom protocol.

@@

From an user experience point of view, we designed our collaboration application around the concept of group conferences (or conference rooms).
Users can create group conferences and invite other users to participate in these, either by adding the other user or by sharing a \gls{URL} (invitation link).
In order to meet user expectation, a group conference provides functionality similar to those found in established applications such as \emph{Skype} or \emph{Google Hangouts}.
Among other common features, one can view a small video of each participant or a large video of a selected participant.
Each participant may use more than one camera, in order to show different angles or objects, e.g. a white board and himself.
The source of the large video can be selected manually or automatically assigned to the participant talking.
Participating users may choose to view only, share their microphone, camera or screen.

Besides the videos, a group conference includes a chat, collaborative text editor, list of users and document sharing.
The major differentiating feature is a time-line bar, for navigating in time.
While the group conference is taking place, this time bar is continuously progressing as time goes by.
The user is free to slide over to any instant in the past or back to real-time.
Changing the instant of time updates the state of not only the video but of other collaboration elements.

Either in real-time or later, users can create bookmarks, displayed in the time-line, for easier navigation in time.
Users can also add content to superimpose on the video.
This content can be as simple as subtitles or as complex as a \gls{HTML} document or JavaScript code to perform an animation.
All the content is searchable, allowing for time navigation by content.

From an early stage, we decided to target the browser platform, due to its ubiquity and breadth of devices, from desktops, to tablets and smartphones.
The nature of web browser applications already follows the hypermedia concept, which makes \emph{WebRTC} the ideal technology to apply the hyper-linked communications concepts.
The advent of \gls{WebRTC}, currently available on the major web browsers, makes it possible to develop video conference web applications without plugins.
However, mobile devices have limited resources for handling a large number of simultaneous video stream.

In order to limit the processing requirement in the client application (browser), we decided to use a server to combine the audio and video streams from all the participants into a single video and a single audio track.
This also helps conserve battery on mobile devices.
Each user listens to the combined audio track and may choose to see a single participant or all of them, in this case using the combined video track.

We realized that it would be useful to add some content, such as bookmarks, an agenda or links to referenced documents, while participating in a conference.
We also realized it would be difficult for the same person to do it while performing a presentation.
As such we devised a method, where the user displays \gls{QR} codes in order to ease content creation.

Figure~\ref{fig:modules} presents the structure of our system, which was divided into six modules:

@@

\paragraph{Application module} responsible for providing information about the relevant modules (\emph{NAT Traversal} and \emph{Signaling}) and user interface to the \emph{Client} in the form of web pages in \gls{HTML} and \emph{JavaScript} libraries through \gls{HTTP}.

\paragraph{Signaling module} responsible for \emph{Client} and \emph{Stream} coordination which will be performed using \emph{WebSockets}.

\paragraph{NAT Traversal module} \gls{STUN} and \gls{TURN} techniques (part of \gls{ICE}) are used by the \emph{Client} and \emph{Stream} modules during the \emph{Signaling} phase that ends with the establishment of the connection between them.

\paragraph{Stream module} responsible to deliver and receive multimedia content from the \emph{Client} using \gls{WebRTC}. This is where audio and video are combined into single streams. It is also here that \gls{QR} codes present in the client streams are detected.

\paragraph{Storage module} provides two main functionalities: store the model information and media recorded. This is the single module responsible for persistent storage. It stores user and communication data as well as all the data required among user sessions. It is also use to store all the communication streams, so that they can be viewed later.

\paragraph{Client module} responsible for the interaction with the user. Runs on the browser on the user device.

Figure \ref{fig:infrastructure} depicts the architecture of our system, being composed by: web server, streaming server (\gls{KMS}), signaling server (\gls{ICE} server), database (MongoDB) and video repository (Kurento Repository).

\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{figures/infrastructure.pdf}
	\caption{System Infrastructure}
        \label{fig:infrastructure}
\end{figure}

In order to simplify our solution, the \emph{Application}, \emph{Signaling}, \emph{Stream} and \emph{Storage} modules are deployed within the same  server, so to facilitate deployment.
To this set of modules, we call \emph{backend}.
The \emph{Application} module is split between client and server part.
The \emph{Signaling} module is implemented together with the application server.

(Application) clients can only access the web server, all other servers are in a firewalled private network.
Access to our streaming servers is given to clients after concluding the signaling phase.
This signaling phase may or not proceed in function of the client access permissions to the particular group conference.

With the use of a centralized architecture, each client only has one \emph{PeerConnection} to our streaming server and content shown to them is changed on demand, either being it an individual or a composite view. 
A peer-to-peer approach would have required each peer to maintain a connection to each of the other participants an one to the streaming server, for recording.
This would seriously limit the size of each group conference by placing high bandwidth and processing requirements on the clients. 
The same can be concluded about instant message delivery.
Each client has only one \emph{WebSocket} connection to the application server which consequently relays the messages to other users.

Our architecture can be scaled by having conference rooms distributed across different servers.
However, all of a group conference participants must be connected to the same server.

\section{Implementation}
\label{chapter:implementation}

Our prototype was built using \gls{KMS} as our media streamer, as \gls{KMS} supports \gls{WebRTC}.
The server side of the \emph{Application} and \emph{Signaling} modules are implemented as a server side web application. 
Our server could be implemented with JavaScript (e.g. with \emph{NodeJS}) or \emph{Java} due to the fact \gls{KMS} provides clients for both technologies. 
We decided to implement our web server using the \emph{PlayFramework}\footnote{\url{https://www.playframework.com/}(acessed March 25, 2016)} using \emph{Java}.
For the collaborative text editor, we resorted to \emph{OT.js}\footnote{\url{http://operational-transformation.github.io}(accessed March 10, 2016)} due to its server and storage implementation choice independence.

By default, \emph{Kurento Repository}, used for storing the recorded video and audio streams, is implemented over \emph{MongoDB}.
For convenience, our storage model also uses the same database.
In order to offer all the functionality, some information about objects must be persistent such as users, groups, relations among users, group memberships, messages, hyper-content, recordings and collaborative editor state.

For the \emph{NAT Traversal} module a public \gls{STUN} server was used for testing our solution.
Nevertheless, we recognize that for a production environment we would need to maintain our own \gls{TURN} servers in order to ensure connectivity to all clients.

Table~\ref{table:apparch} presents the client application architecture and the underlying technologies used.
Both \emph{Mozilla Firefox} and \emph{Google Chrome} browsers are supported.
\emph{Adapter.js} and \emph{jQuery} ensure that our application is compatible with the most popular web browsers.
\emph{Bootstrap} is used to make the user interface more appellative and responsive, adapting to mobile devices with different screen sizes.
\emph{OT.js} is used to build the collaborative text editor.

\begin{table}
\centering
	\caption{Application Architecture}
	\label{table:apparch}
